---
title: "Stats Integration"
author: "Enrique Chimal"
date: "2026-01-13"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Basic Settings

```{r}
setwd("/N/u/echimal/Quartz/Desktop/CLR_MRI/Human_GeoMx_Sep2025/SpaceJam_results")
library(glmmTMB)
library(dplyr)
library(emmeans)
library(broom.mixed)
library(ggplot2)
library(pheatmap)
```

### Load Dataset

```{r}
df <- read.csv("spatial_celltype_proportions.csv")

df$disease_status <- factor(df$disease_status, levels = c("Control", "AD-CAA"))
df$Scan_ID <- factor(df$Scan_ID)
df$celltype <- factor(df$celltype)
df$region <- factor(df$region)
df$pathology <- factor(df$pathology)

```

### Beta Mixed Model setup

```{r}
#Fix 0/1 values (for beta regression)
eps <- 1e-6
df$rel_abundance <- pmin(pmax(df$rel_abundance, eps), 1 - eps)

```

# Arteries Subset

```{r}
########=======
arteries <- df %>% filter(region == "Arteries")

# Build a 3-level "group":
# Control only has AmyloidFree -> Define the label explicitly
arteries <- arteries %>%
  mutate(
    group = case_when(
      disease_status == "Control" ~ "Control_AmyloidFree",
      disease_status == "AD-CAA" & pathology == "AmyloidFree" ~ "AD_AmyloidFree",
      disease_status == "AD-CAA" & pathology == "Amyloid" ~ "AD_Amyloid",
      TRUE ~ NA_character_
    ),
    group = factor(group, levels = c("Control_AmyloidFree", "AD_AmyloidFree", "AD_Amyloid"))
  ) %>%
  filter(!is.na(group))

# Compute AD amyloid weight (within arteries)
# w = proportion of AD arterial ROIs that are Amyloid
w <- arteries %>%
  filter(disease_status == "AD-CAA") %>%
  summarise(w = mean(pathology == "Amyloid")) %>%
  pull(w)

cat("AD amyloid weight w =", w, "\n")

#Fit per celltype beta mixed model + extract 3 contrasts 
contrast_results <- list()

for (ct in levels(arteries$celltype)) {
  
  tmp <- arteries %>% filter(celltype == ct)
  
  # basic sanity
  if (n_distinct(tmp$Scan_ID) < 2) next
  if (n_distinct(tmp$group) < 2) next
  
  fit <- glmmTMB(
    rel_abundance ~ group + (1 | Scan_ID),
    data = tmp,
    family = beta_family(link = "logit")
  )
  
  em_link <- emmeans(fit, ~ group)  # link scale (logit)
  
  # Contrasts on link scale:
  # Disease_effect: AD_AmyloidFree - Control_AmyloidFree
  # Amyloid_effect: AD_Amyloid - AD_AmyloidFree
  # AD_overall_vs_Control: ( (1-w)*AD_AmyloidFree + w*AD_Amyloid ) - Control_AmyloidFree
  contr <- contrast(
    em_link,
    method = list(
      Disease_effect        = c(-1,  1,  0),
      Amyloid_effect        = c( 0, -1,  1),
      AD_overall_vs_Control = c(-1, (1 - w), w)
    )
  )
  
  out <- as.data.frame(summary(contr, infer = TRUE)) %>%
    mutate(celltype = ct)
  
  contrast_results[[ct]] <- out
}

contrasts3 <- bind_rows(contrast_results)

#Multiple Hypothesis Testing
contrasts3 <- contrasts3 %>%
  group_by(contrast) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()

write.csv(contrasts3, "arteries_3contrasts_betaMM.csv", row.names = FALSE)

sig_counts <- contrasts3 %>%
  group_by(contrast) %>%
  summarise(
    n_tested = n(),
    n_sig_FDR_0.05 = sum(p_adj < 0.05, na.rm = TRUE),
    .groups = "drop"
  )

print(sig_counts)

#HeatMap
heatmap_df <- contrasts3 %>%
  select(celltype, contrast, estimate, p_adj) %>%
  tidyr::pivot_wider(
    names_from = contrast,
    values_from = c(estimate, p_adj),
    names_sep = "___"
  )

# Extract effect matrix (3 columns)
effect_cols <- c("estimate___Disease_effect",
                 "estimate___Amyloid_effect",
                 "estimate___AD_overall_vs_Control")

padj_cols <- c("p_adj___Disease_effect",
               "p_adj___Amyloid_effect",
               "p_adj___AD_overall_vs_Control")

heatmap_mat <- as.matrix(heatmap_df[, effect_cols])
rownames(heatmap_mat) <- heatmap_df$celltype
colnames(heatmap_mat) <- c("Disease_effect", "Amyloid_effect", "AD_overall_vs_Control")

# Clean non-finite values for clustering safety
heatmap_mat_clean <- heatmap_mat
heatmap_mat_clean[!is.finite(heatmap_mat_clean)] <- 0

# Optional: add significance stars as annotation labels
padj_mat <- as.matrix(heatmap_df[, padj_cols])
rownames(padj_mat) <- heatmap_df$celltype
colnames(padj_mat) <- c("Disease_effect", "Amyloid_effect", "AD_overall_vs_Control")

sig_labels <- matrix("", nrow = nrow(padj_mat), ncol = ncol(padj_mat),
                     dimnames = dimnames(padj_mat))
sig_labels[padj_mat < 0.05] <- "*"
sig_labels[padj_mat < 0.01] <- "**"
sig_labels[padj_mat < 0.001] <- "***"
sig_labels[is.na(padj_mat)] <- ""

pheatmap(
  heatmap_mat_clean,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  scale = "none",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  border_color = NA,
  fontsize_row = 8,
  fontsize_col = 10,
  display_numbers = sig_labels,
  number_color = "black",
  main = "Arterial Cell-Type Changes\nDisease vs Amyloid vs Weighted AD Overall"
)
```

# Parenchyma Subset

```{r}
########=======
parenchyma <- df %>% filter(region == "Parenchyma")

# Build a 3-level "group":
# Control only has AmyloidFree -> Define the label explicitly
parenchyma <- parenchyma %>%
  mutate(
    group = case_when(
      disease_status == "Control" ~ "Control_AmyloidFree",
      disease_status == "AD-CAA" & pathology == "AmyloidFree" ~ "AD_AmyloidFree",
      disease_status == "AD-CAA" & pathology == "Amyloid" ~ "AD_Amyloid",
      TRUE ~ NA_character_
    ),
    group = factor(group, levels = c("Control_AmyloidFree", "AD_AmyloidFree", "AD_Amyloid"))
  ) %>%
  filter(!is.na(group))

# Compute AD amyloid weight (within arteries)
# w = proportion of AD arterial ROIs that are Amyloid
w <- parenchyma %>%
  filter(disease_status == "AD-CAA") %>%
  summarise(w = mean(pathology == "Amyloid")) %>%
  pull(w)

cat("AD amyloid weight w =", w, "\n")

#Fit per celltype beta mixed model + extract 3 contrasts 
contrast_results <- list()

for (ct in levels(parenchyma$celltype)) {
  
  tmp <- parenchyma %>% filter(celltype == ct)
  
  # basic sanity
  if (n_distinct(tmp$Scan_ID) < 2) next
  if (n_distinct(tmp$group) < 2) next
  
  fit <- glmmTMB(
    rel_abundance ~ group + (1 | Scan_ID),
    data = tmp,
    family = beta_family(link = "logit")
  )
  
  em_link <- emmeans(fit, ~ group)  # link scale (logit)
  
  # Contrasts on link scale:
  # Disease_effect: AD_AmyloidFree - Control_AmyloidFree
  # Amyloid_effect: AD_Amyloid - AD_AmyloidFree
  # AD_overall_vs_Control: ( (1-w)*AD_AmyloidFree + w*AD_Amyloid ) - Control_AmyloidFree
  contr <- contrast(
    em_link,
    method = list(
      Disease_effect        = c(-1,  1,  0),
      Amyloid_effect        = c( 0, -1,  1),
      AD_overall_vs_Control = c(-1, (1 - w), w)
    )
  )
  
  out <- as.data.frame(summary(contr, infer = TRUE)) %>%
    mutate(celltype = ct)
  
  contrast_results[[ct]] <- out
}

contrasts3 <- bind_rows(contrast_results)

#Multiple Hypothesis Testing
contrasts3 <- contrasts3 %>%
  group_by(contrast) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()

write.csv(contrasts3, "parenchyma_3contrasts_betaMM.csv", row.names = FALSE)

sig_counts <- contrasts3 %>%
  group_by(contrast) %>%
  summarise(
    n_tested = n(),
    n_sig_FDR_0.05 = sum(p_adj < 0.05, na.rm = TRUE),
    .groups = "drop"
  )

print(sig_counts)

#HeatMap
heatmap_df <- contrasts3 %>%
  select(celltype, contrast, estimate, p_adj) %>%
  tidyr::pivot_wider(
    names_from = contrast,
    values_from = c(estimate, p_adj),
    names_sep = "___"
  )

# Extract effect matrix (3 columns)
effect_cols <- c("estimate___Disease_effect",
                 "estimate___Amyloid_effect",
                 "estimate___AD_overall_vs_Control")

padj_cols <- c("p_adj___Disease_effect",
               "p_adj___Amyloid_effect",
               "p_adj___AD_overall_vs_Control")

heatmap_mat <- as.matrix(heatmap_df[, effect_cols])
rownames(heatmap_mat) <- heatmap_df$celltype
colnames(heatmap_mat) <- c("Disease_effect", "Amyloid_effect", "AD_overall_vs_Control")

# Clean non-finite values for clustering safety
heatmap_mat_clean <- heatmap_mat
heatmap_mat_clean[!is.finite(heatmap_mat_clean)] <- 0

# Optional: add significance stars as annotation labels
padj_mat <- as.matrix(heatmap_df[, padj_cols])
rownames(padj_mat) <- heatmap_df$celltype
colnames(padj_mat) <- c("Disease_effect", "Amyloid_effect", "AD_overall_vs_Control")

sig_labels <- matrix("", nrow = nrow(padj_mat), ncol = ncol(padj_mat),
                     dimnames = dimnames(padj_mat))
sig_labels[padj_mat < 0.05] <- "*"
sig_labels[padj_mat < 0.01] <- "**"
sig_labels[padj_mat < 0.001] <- "***"
sig_labels[is.na(padj_mat)] <- ""

pheatmap(
  heatmap_mat_clean,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  scale = "none",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  border_color = NA,
  fontsize_row = 8,
  fontsize_col = 10,
  display_numbers = sig_labels,
  number_color = "black",
  main = "Parenchyma Cell-Type Changes\nDisease vs Amyloid vs Weighted AD Overall"
)
```

# Capillaries Subset

```{r}
########=======
capillaries <- df %>% filter(region == "Capillaries")

# Build a 3-level "group":
# Control only has AmyloidFree -> Define the label explicitly
capillaries <- capillaries %>%
  mutate(
    group = case_when(
      disease_status == "Control" ~ "Control_AmyloidFree",
      disease_status == "AD-CAA" & pathology == "AmyloidFree" ~ "AD_AmyloidFree",
      disease_status == "AD-CAA" & pathology == "Amyloid" ~ "AD_Amyloid",
      TRUE ~ NA_character_
    ),
    group = factor(group, levels = c("Control_AmyloidFree", "AD_AmyloidFree", "AD_Amyloid"))
  ) %>%
  filter(!is.na(group))

# Compute AD amyloid weight (within arteries)
# w = proportion of AD arterial ROIs that are Amyloid
w <- capillaries %>%
  filter(disease_status == "AD-CAA") %>%
  summarise(w = mean(pathology == "Amyloid")) %>%
  pull(w)

cat("AD amyloid weight w =", w, "\n")

#Fit per celltype beta mixed model + extract 3 contrasts 
contrast_results <- list()

for (ct in levels(capillaries$celltype)) {
  
  tmp <- capillaries %>% filter(celltype == ct)
  
  # basic sanity
  if (n_distinct(tmp$Scan_ID) < 2) next
  if (n_distinct(tmp$group) < 2) next
  
  fit <- glmmTMB(
    rel_abundance ~ group + (1 | Scan_ID),
    data = tmp,
    family = beta_family(link = "logit")
  )
  
  em_link <- emmeans(fit, ~ group)  # link scale (logit)
  
  # Contrasts on link scale:
  # Disease_effect: AD_AmyloidFree - Control_AmyloidFree
  # Amyloid_effect: AD_Amyloid - AD_AmyloidFree
  # AD_overall_vs_Control: ( (1-w)*AD_AmyloidFree + w*AD_Amyloid ) - Control_AmyloidFree
  contr <- contrast(
    em_link,
    method = list(
      Disease_effect        = c(-1,  1,  0),
      Amyloid_effect        = c( 0, -1,  1),
      AD_overall_vs_Control = c(-1, (1 - w), w)
    )
  )
  
  out <- as.data.frame(summary(contr, infer = TRUE)) %>%
    mutate(celltype = ct)
  
  contrast_results[[ct]] <- out
}

contrasts3 <- bind_rows(contrast_results)

#Multiple Hypothesis Testing
contrasts3 <- contrasts3 %>%
  group_by(contrast) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()

write.csv(contrasts3, "capillaries_3contrasts_betaMM.csv", row.names = FALSE)

sig_counts <- contrasts3 %>%
  group_by(contrast) %>%
  summarise(
    n_tested = n(),
    n_sig_FDR_0.05 = sum(p_adj < 0.05, na.rm = TRUE),
    .groups = "drop"
  )

print(sig_counts)

#HeatMap
heatmap_df <- contrasts3 %>%
  select(celltype, contrast, estimate, p_adj) %>%
  tidyr::pivot_wider(
    names_from = contrast,
    values_from = c(estimate, p_adj),
    names_sep = "___"
  )

# Extract effect matrix (3 columns)
effect_cols <- c("estimate___Disease_effect",
                 "estimate___Amyloid_effect",
                 "estimate___AD_overall_vs_Control")

padj_cols <- c("p_adj___Disease_effect",
               "p_adj___Amyloid_effect",
               "p_adj___AD_overall_vs_Control")

heatmap_mat <- as.matrix(heatmap_df[, effect_cols])
rownames(heatmap_mat) <- heatmap_df$celltype
colnames(heatmap_mat) <- c("Disease_effect", "Amyloid_effect", "AD_overall_vs_Control")

# Clean non-finite values for clustering safety
heatmap_mat_clean <- heatmap_mat
heatmap_mat_clean[!is.finite(heatmap_mat_clean)] <- 0

# Optional: add significance stars as annotation labels
padj_mat <- as.matrix(heatmap_df[, padj_cols])
rownames(padj_mat) <- heatmap_df$celltype
colnames(padj_mat) <- c("Disease_effect", "Amyloid_effect", "AD_overall_vs_Control")

sig_labels <- matrix("", nrow = nrow(padj_mat), ncol = ncol(padj_mat),
                     dimnames = dimnames(padj_mat))
sig_labels[padj_mat < 0.05] <- "*"
sig_labels[padj_mat < 0.01] <- "**"
sig_labels[padj_mat < 0.001] <- "***"
sig_labels[is.na(padj_mat)] <- ""

pheatmap(
  heatmap_mat_clean,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  scale = "none",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  border_color = NA,
  fontsize_row = 8,
  fontsize_col = 10,
  display_numbers = sig_labels,
  number_color = "black",
  main = "Capillaries Cell-Type Changes\nDisease vs Amyloid vs Weighted AD Overall"
)
```

## Sanity Check with Only Parenchyma & Amyloid Effect (to recapitulate AD Paper 2026)

```{r}
parenchyma <- df %>% 
  filter(region == "Parenchyma",
         disease_status == "AD-CAA")
results_amyloid_par <- list()

for (ct in levels(parenchyma$celltype)) {
  
  tmp <- parenchyma %>% filter(celltype == ct)
  if (n_distinct(tmp$Scan_ID) < 2) next
  if (n_distinct(tmp$pathology) < 2) next
  
  fit <- glmmTMB(
    rel_abundance ~ pathology + (1 | Scan_ID),
    data = tmp,
    family = beta_family(link = "logit")
  )
  
  emm <- emmeans(fit, ~ pathology, type = "response")
  cont <- contrast(emm, method = "revpairwise")
  
  out <- as.data.frame(summary(cont, infer = TRUE))
  out$celltype <- ct
  
  results_amyloid_par[[ct]] <- out
}

amyloid_par_df <- bind_rows(results_amyloid_par) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH"))

sig_amyloid_par <- amyloid_par_df %>%
  filter(p_adj < 0.05)

n_sig <- n_distinct(sig_amyloid_par$celltype)
n_total <- n_distinct(amyloid_par_df$celltype)

n_sig
```

```{r}
sig_cts <- sig_amyloid_par$celltype
sig_cts

plot_df <- df %>%
  filter(
    region == "Parenchyma",
    disease_status == "AD-CAA",
    celltype %in% sig_cts
  ) %>%
  mutate(
    pathology = factor(pathology, levels = c("AmyloidFree", "Amyloid"))
  )

library(ggplot2)

ggplot(plot_df,
       aes(x = pathology,
           y = rel_abundance,
           color = pathology)) +
  geom_boxplot(
    outlier.shape = NA,
    width = 0.5,
    alpha = 0.6
  ) +
  geom_jitter(
    width = 0.15,
    size = 1.8,
    alpha = 0.6
  ) +
  facet_wrap(~ celltype, scales = "free_y", ncol = 2) +
  scale_color_manual(
    values = c("AmyloidFree" = "#2C7BB6",
               "Amyloid" = "#D7191C")
  ) +
  labs(
    title = "Parenchymal Amyloid-Associated Cell-Type Changes (AD-CAA)",
    subtitle = "Only cell types significant after FDR correction",
    x = "",
    y = "Relative abundance"
  ) +
  theme_classic(base_size = 13) +
  theme(
    legend.position = "none",
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

### Gene Enrichment

```{r}
#From the Seurat Code for snRNA:
markers <- read.table("AD_CAA.ctx.seurat_withsubsets_clusterDEGs.txt")
```

```{r}
library(dplyr)

topN <- 300  # 30 to 100 is typical

marker_lists <- markers %>%
  filter(cluster %in% sig_cts) %>%
  group_by(cluster) %>%
  arrange(p_val_adj, desc(avg_log2FC)) %>%
  slice_head(n = topN) %>%
  summarise(
    celltype = first(cluster),
    genes = list(unique(gene)),
    .groups = "drop"
  )

marker_lists
sapply(marker_lists$genes, length)

```

```{r}
dir.create("enrichment_inputs", showWarnings = FALSE)

for (i in seq_len(nrow(marker_lists))) {
  ct <- marker_lists$celltype[i]
  write.table(marker_lists$genes[[i]],
              file = file.path("enrichment_inputs", paste0(ct, "_top", topN, "_genes.txt")),
              row.names = FALSE, col.names = FALSE, quote = FALSE)
}
```

Running the top 300 genes on Reactome OPC1 recapitulates the Matrisome pathways that are described on the 2026 Chimal et al paper.

Further Statistics need to be discussed and agreed to align with the objectives of the paper.
