---
title: "Stats Integration"
author: "Enrique Chimal"
date: "2026-01-13"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)



```

### Basic Settings

```{r}
setwd("/N/u/echimal/Quartz/Desktop/CLR_MRI/Human_GeoMx_Sep2025/SpaceJam_results")
library(glmmTMB)
library(dplyr)
library(emmeans)
library(broom.mixed)
library(ggplot2)
library(pheatmap)
library(dplyr)
library(tidyr)
```

### Load Dataset

```{r}
df <- read.csv("spatial_celltype_proportions.csv")

df$disease_status <- factor(df$disease_status, levels = c("Control", "AD-CAA"))
df$Scan_ID <- factor(df$Scan_ID)
df$celltype <- factor(df$celltype)
df$region <- factor(df$region)
df$pathology <- factor(df$pathology)

```

### Beta Mixed Model setup

```{r}
#Fix 0/1 values (for beta regression)
eps <- 1e-6
df$rel_abundance <- pmin(pmax(df$rel_abundance, eps), 1 - eps)

```

# Arteries Subset

```{r}
########=======
arteries <- df %>% filter(region == "Arteries")

# Build a 3-level "group":
# Control only has AmyloidFree -> Define the label explicitly
arteries <- arteries %>%
  mutate(
    group = case_when(
      disease_status == "Control" ~ "Control_AmyloidFree",
      disease_status == "AD-CAA" & pathology == "AmyloidFree" ~ "AD_AmyloidFree",
      disease_status == "AD-CAA" & pathology == "Amyloid" ~ "AD_Amyloid",
      TRUE ~ NA_character_
    ),
    group = factor(group, levels = c("Control_AmyloidFree", "AD_AmyloidFree", "AD_Amyloid"))
  ) %>%
  filter(!is.na(group))

# Compute AD amyloid weight (within arteries)
# w = proportion of AD arterial ROIs that are Amyloid
w <- arteries %>%
  filter(disease_status == "AD-CAA") %>%
  summarise(w = mean(pathology == "Amyloid")) %>%
  pull(w)

cat("AD amyloid weight w =", w, "\n")

#Fit per celltype beta mixed model + extract 3 contrasts 
contrast_results <- list()

for (ct in levels(arteries$celltype)) {
  
  tmp <- arteries %>% filter(celltype == ct)
  
  # basic sanity
  if (n_distinct(tmp$Scan_ID) < 2) next
  if (n_distinct(tmp$group) < 2) next
  
  fit <- glmmTMB(
    rel_abundance ~ group + (1 | Scan_ID),
    data = tmp,
    family = beta_family(link = "logit")
  )
  
  em_link <- emmeans(fit, ~ group)  # link scale (logit)
  
  # Contrasts on link scale:
  # Disease_effect: AD_AmyloidFree - Control_AmyloidFree
  # Amyloid_effect: AD_Amyloid - AD_AmyloidFree
  # AD_overall_vs_Control: ( (1-w)*AD_AmyloidFree + w*AD_Amyloid ) - Control_AmyloidFree
  contr <- contrast(
    em_link,
    method = list(
      Disease_effect        = c(-1,  1,  0),
      Amyloid_effect        = c( 0, -1,  1),
      AD_overall_vs_Control = c(-1, (1 - w), w)
    )
  )
  
  out <- as.data.frame(summary(contr, infer = TRUE)) %>%
    mutate(celltype = ct)
  
  contrast_results[[ct]] <- out
}

contrasts3 <- bind_rows(contrast_results)

#Multiple Hypothesis Testing
contrasts3 <- contrasts3 %>%
  group_by(contrast) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()

write.csv(contrasts3, "arteries_3contrasts_betaMM.csv", row.names = FALSE)

sig_counts <- contrasts3 %>%
  group_by(contrast) %>%
  summarise(
    n_tested = n(),
    n_sig_FDR_0.05 = sum(p_adj < 0.05, na.rm = TRUE),
    .groups = "drop"
  )

print(sig_counts)

#HeatMap
heatmap_df <- contrasts3 %>%
  select(celltype, contrast, estimate, p_adj) %>%
  tidyr::pivot_wider(
    names_from = contrast,
    values_from = c(estimate, p_adj),
    names_sep = "___"
  )

# Extract effect matrix (3 columns)
effect_cols <- c("estimate___Disease_effect",
                 "estimate___Amyloid_effect",
                 "estimate___AD_overall_vs_Control")

padj_cols <- c("p_adj___Disease_effect",
               "p_adj___Amyloid_effect",
               "p_adj___AD_overall_vs_Control")

heatmap_mat <- as.matrix(heatmap_df[, effect_cols])
rownames(heatmap_mat) <- heatmap_df$celltype
colnames(heatmap_mat) <- c("Disease_effect", "Amyloid_effect", "AD_overall_vs_Control")

# Clean non-finite values for clustering safety
heatmap_mat_clean <- heatmap_mat
heatmap_mat_clean[!is.finite(heatmap_mat_clean)] <- 0

# Optional: add significance stars as annotation labels
padj_mat <- as.matrix(heatmap_df[, padj_cols])
rownames(padj_mat) <- heatmap_df$celltype
colnames(padj_mat) <- c("Disease_effect", "Amyloid_effect", "AD_overall_vs_Control")

sig_labels <- matrix("", nrow = nrow(padj_mat), ncol = ncol(padj_mat),
                     dimnames = dimnames(padj_mat))
sig_labels[padj_mat < 0.05] <- "*"
sig_labels[padj_mat < 0.01] <- "**"
sig_labels[padj_mat < 0.001] <- "***"
sig_labels[is.na(padj_mat)] <- ""

pheatmap(
  heatmap_mat_clean,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  scale = "none",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  border_color = NA,
  fontsize_row = 8,
  fontsize_col = 10,
  display_numbers = sig_labels,
  number_color = "black",
  main = "Arterial Cell-Type Changes\nDisease vs Amyloid vs Weighted AD Overall"
)
```

# Parenchyma Subset

```{r}
########=======
parenchyma <- df %>% filter(region == "Parenchyma")

# Build a 3-level "group":
# Control only has AmyloidFree -> Define the label explicitly
parenchyma <- parenchyma %>%
  mutate(
    group = case_when(
      disease_status == "Control" ~ "Control_AmyloidFree",
      disease_status == "AD-CAA" & pathology == "AmyloidFree" ~ "AD_AmyloidFree",
      disease_status == "AD-CAA" & pathology == "Amyloid" ~ "AD_Amyloid",
      TRUE ~ NA_character_
    ),
    group = factor(group, levels = c("Control_AmyloidFree", "AD_AmyloidFree", "AD_Amyloid"))
  ) %>%
  filter(!is.na(group))

# Compute AD amyloid weight (within arteries)
# w = proportion of AD arterial ROIs that are Amyloid
w <- parenchyma %>%
  filter(disease_status == "AD-CAA") %>%
  summarise(w = mean(pathology == "Amyloid")) %>%
  pull(w)

cat("AD amyloid weight w =", w, "\n")

#Fit per celltype beta mixed model + extract 3 contrasts 
contrast_results <- list()

for (ct in levels(parenchyma$celltype)) {
  
  tmp <- parenchyma %>% filter(celltype == ct)
  
  # basic sanity
  if (n_distinct(tmp$Scan_ID) < 2) next
  if (n_distinct(tmp$group) < 2) next
  
  fit <- glmmTMB(
    rel_abundance ~ group + (1 | Scan_ID),
    data = tmp,
    family = beta_family(link = "logit")
  )
  
  em_link <- emmeans(fit, ~ group)  # link scale (logit)
  
  # Contrasts on link scale:
  # Disease_effect: AD_AmyloidFree - Control_AmyloidFree
  # Amyloid_effect: AD_Amyloid - AD_AmyloidFree
  # AD_overall_vs_Control: ( (1-w)*AD_AmyloidFree + w*AD_Amyloid ) - Control_AmyloidFree
  contr <- contrast(
    em_link,
    method = list(
      Disease_effect        = c(-1,  1,  0),
      Amyloid_effect        = c( 0, -1,  1),
      AD_overall_vs_Control = c(-1, (1 - w), w)
    )
  )
  
  out <- as.data.frame(summary(contr, infer = TRUE)) %>%
    mutate(celltype = ct)
  
  contrast_results[[ct]] <- out
}

contrasts3 <- bind_rows(contrast_results)

#Multiple Hypothesis Testing
contrasts3 <- contrasts3 %>%
  group_by(contrast) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()

write.csv(contrasts3, "parenchyma_3contrasts_betaMM.csv", row.names = FALSE)

sig_counts <- contrasts3 %>%
  group_by(contrast) %>%
  summarise(
    n_tested = n(),
    n_sig_FDR_0.05 = sum(p_adj < 0.05, na.rm = TRUE),
    .groups = "drop"
  )

print(sig_counts)

#HeatMap
heatmap_df <- contrasts3 %>%
  select(celltype, contrast, estimate, p_adj) %>%
  tidyr::pivot_wider(
    names_from = contrast,
    values_from = c(estimate, p_adj),
    names_sep = "___"
  )

# Extract effect matrix (3 columns)
effect_cols <- c("estimate___Disease_effect",
                 "estimate___Amyloid_effect",
                 "estimate___AD_overall_vs_Control")

padj_cols <- c("p_adj___Disease_effect",
               "p_adj___Amyloid_effect",
               "p_adj___AD_overall_vs_Control")

heatmap_mat <- as.matrix(heatmap_df[, effect_cols])
rownames(heatmap_mat) <- heatmap_df$celltype
colnames(heatmap_mat) <- c("Disease_effect", "Amyloid_effect", "AD_overall_vs_Control")

# Clean non-finite values for clustering safety
heatmap_mat_clean <- heatmap_mat
heatmap_mat_clean[!is.finite(heatmap_mat_clean)] <- 0

# Optional: add significance stars as annotation labels
padj_mat <- as.matrix(heatmap_df[, padj_cols])
rownames(padj_mat) <- heatmap_df$celltype
colnames(padj_mat) <- c("Disease_effect", "Amyloid_effect", "AD_overall_vs_Control")

sig_labels <- matrix("", nrow = nrow(padj_mat), ncol = ncol(padj_mat),
                     dimnames = dimnames(padj_mat))
sig_labels[padj_mat < 0.05] <- "*"
sig_labels[padj_mat < 0.01] <- "**"
sig_labels[padj_mat < 0.001] <- "***"
sig_labels[is.na(padj_mat)] <- ""

pheatmap(
  heatmap_mat_clean,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  scale = "none",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  border_color = NA,
  fontsize_row = 8,
  fontsize_col = 10,
  display_numbers = sig_labels,
  number_color = "black",
  main = "Parenchyma Cell-Type Changes\nDisease vs Amyloid vs Weighted AD Overall"
)
```

# Capillaries Subset

```{r}
########=======
capillaries <- df %>% filter(region == "Capillaries")

# Build a 3-level "group":
# Control only has AmyloidFree -> Define the label explicitly
capillaries <- capillaries %>%
  mutate(
    group = case_when(
      disease_status == "Control" ~ "Control_AmyloidFree",
      disease_status == "AD-CAA" & pathology == "AmyloidFree" ~ "AD_AmyloidFree",
      disease_status == "AD-CAA" & pathology == "Amyloid" ~ "AD_Amyloid",
      TRUE ~ NA_character_
    ),
    group = factor(group, levels = c("Control_AmyloidFree", "AD_AmyloidFree", "AD_Amyloid"))
  ) %>%
  filter(!is.na(group))

# Compute AD amyloid weight (within arteries)
# w = proportion of AD arterial ROIs that are Amyloid
w <- capillaries %>%
  filter(disease_status == "AD-CAA") %>%
  summarise(w = mean(pathology == "Amyloid")) %>%
  pull(w)

cat("AD amyloid weight w =", w, "\n")

#Fit per celltype beta mixed model + extract 3 contrasts 
contrast_results <- list()

for (ct in levels(capillaries$celltype)) {
  
  tmp <- capillaries %>% filter(celltype == ct)
  
  # basic sanity
  if (n_distinct(tmp$Scan_ID) < 2) next
  if (n_distinct(tmp$group) < 2) next
  
  fit <- glmmTMB(
    rel_abundance ~ group + (1 | Scan_ID),
    data = tmp,
    family = beta_family(link = "logit")
  )
  
  em_link <- emmeans(fit, ~ group)  # link scale (logit)
  
  # Contrasts on link scale:
  # Disease_effect: AD_AmyloidFree - Control_AmyloidFree
  # Amyloid_effect: AD_Amyloid - AD_AmyloidFree
  # AD_overall_vs_Control: ( (1-w)*AD_AmyloidFree + w*AD_Amyloid ) - Control_AmyloidFree
  contr <- contrast(
    em_link,
    method = list(
      Disease_effect        = c(-1,  1,  0),
      Amyloid_effect        = c( 0, -1,  1),
      AD_overall_vs_Control = c(-1, (1 - w), w)
    )
  )
  
  out <- as.data.frame(summary(contr, infer = TRUE)) %>%
    mutate(celltype = ct)
  
  contrast_results[[ct]] <- out
}

contrasts3 <- bind_rows(contrast_results)

#Multiple Hypothesis Testing
contrasts3 <- contrasts3 %>%
  group_by(contrast) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()

write.csv(contrasts3, "capillaries_3contrasts_betaMM.csv", row.names = FALSE)

sig_counts <- contrasts3 %>%
  group_by(contrast) %>%
  summarise(
    n_tested = n(),
    n_sig_FDR_0.05 = sum(p_adj < 0.05, na.rm = TRUE),
    .groups = "drop"
  )

print(sig_counts)

#HeatMap
heatmap_df <- contrasts3 %>%
  select(celltype, contrast, estimate, p_adj) %>%
  tidyr::pivot_wider(
    names_from = contrast,
    values_from = c(estimate, p_adj),
    names_sep = "___"
  )

# Extract effect matrix (3 columns)
effect_cols <- c("estimate___Disease_effect",
                 "estimate___Amyloid_effect",
                 "estimate___AD_overall_vs_Control")

padj_cols <- c("p_adj___Disease_effect",
               "p_adj___Amyloid_effect",
               "p_adj___AD_overall_vs_Control")

heatmap_mat <- as.matrix(heatmap_df[, effect_cols])
rownames(heatmap_mat) <- heatmap_df$celltype
colnames(heatmap_mat) <- c("Disease_effect", "Amyloid_effect", "AD_overall_vs_Control")

# Clean non-finite values for clustering safety
heatmap_mat_clean <- heatmap_mat
heatmap_mat_clean[!is.finite(heatmap_mat_clean)] <- 0

# Optional: add significance stars as annotation labels
padj_mat <- as.matrix(heatmap_df[, padj_cols])
rownames(padj_mat) <- heatmap_df$celltype
colnames(padj_mat) <- c("Disease_effect", "Amyloid_effect", "AD_overall_vs_Control")

sig_labels <- matrix("", nrow = nrow(padj_mat), ncol = ncol(padj_mat),
                     dimnames = dimnames(padj_mat))
sig_labels[padj_mat < 0.05] <- "*"
sig_labels[padj_mat < 0.01] <- "**"
sig_labels[padj_mat < 0.001] <- "***"
sig_labels[is.na(padj_mat)] <- ""

pheatmap(
  heatmap_mat_clean,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  scale = "none",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  border_color = NA,
  fontsize_row = 8,
  fontsize_col = 10,
  display_numbers = sig_labels,
  number_color = "black",
  main = "Capillaries Cell-Type Changes\nDisease vs Amyloid vs Weighted AD Overall"
)
```

## Sanity Check with Only Parenchyma & Amyloid Effect (to recapitulate AD Paper 2026)

```{r}
parenchyma <- df %>% 
  filter(region == "Parenchyma",
         disease_status == "AD-CAA")
results_amyloid_par <- list()

for (ct in levels(parenchyma$celltype)) {
  
  tmp <- parenchyma %>% filter(celltype == ct)
  if (n_distinct(tmp$Scan_ID) < 2) next
  if (n_distinct(tmp$pathology) < 2) next
  
  fit <- glmmTMB(
    rel_abundance ~ pathology + (1 | Scan_ID),
    data = tmp,
    family = beta_family(link = "logit")
  )
  
  emm <- emmeans(fit, ~ pathology, type = "response")
  cont <- contrast(emm, method = "revpairwise")
  
  out <- as.data.frame(summary(cont, infer = TRUE))
  out$celltype <- ct
  
  results_amyloid_par[[ct]] <- out
}

amyloid_par_df <- bind_rows(results_amyloid_par) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH"))

sig_amyloid_par <- amyloid_par_df %>%
  filter(p_adj < 0.05)

n_sig <- n_distinct(sig_amyloid_par$celltype)
n_total <- n_distinct(amyloid_par_df$celltype)

n_sig
```

```{r}
sig_cts <- sig_amyloid_par$celltype
sig_cts

plot_df <- df %>%
  filter(
    region == "Parenchyma",
    disease_status == "AD-CAA",
    celltype %in% sig_cts
  ) %>%
  mutate(
    pathology = factor(pathology, levels = c("AmyloidFree", "Amyloid"))
  )

library(ggplot2)

ggplot(plot_df,
       aes(x = pathology,
           y = rel_abundance,
           color = pathology)) +
  geom_boxplot(
    outlier.shape = NA,
    width = 0.5,
    alpha = 0.6
  ) +
  geom_jitter(
    width = 0.15,
    size = 1.8,
    alpha = 0.6
  ) +
  facet_wrap(~ celltype, scales = "free_y", ncol = 2) +
  scale_color_manual(
    values = c("AmyloidFree" = "#2C7BB6",
               "Amyloid" = "#D7191C")
  ) +
  labs(
    title = "Parenchymal Amyloid-Associated Cell-Type Changes (AD-CAA)",
    subtitle = "Only cell types significant after FDR correction",
    x = "",
    y = "Relative abundance"
  ) +
  theme_classic(base_size = 13) +
  theme(
    legend.position = "none",
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

### Gene Enrichment

```{r}
#From the Seurat Code for snRNA:
markers <- read.table("AD_CAA.ctx.seurat_withsubsets_clusterDEGs.txt")
```

```{r}
library(dplyr)

topN <- 300  # 30 to 100 is typical

marker_lists <- markers %>%
  filter(cluster %in% sig_cts) %>%
  group_by(cluster) %>%
  arrange(p_val_adj, desc(avg_log2FC)) %>%
  slice_head(n = topN) %>%
  summarise(
    celltype = first(cluster),
    genes = list(unique(gene)),
    .groups = "drop"
  )

marker_lists
sapply(marker_lists$genes, length)

```

```{r}
dir.create("enrichment_inputs", showWarnings = FALSE)

for (i in seq_len(nrow(marker_lists))) {
  ct <- marker_lists$celltype[i]
  write.table(marker_lists$genes[[i]],
              file = file.path("enrichment_inputs", paste0(ct, "_top", topN, "_genes.txt")),
              row.names = FALSE, col.names = FALSE, quote = FALSE)
}
```

Running the top 300 genes on Reactome OPC1 recapitulates the Matrisome pathways that are described on the 2026 Chimal et al paper.

Further Statistics need to be discussed and agreed to align with the objectives of the paper.

# Amyloid effect (All regions \~ Only AD_CAA)

```{r}
library(dplyr)

df <- read.csv("spatial_celltype_proportions.csv", stringsAsFactors = FALSE)

# Make sure rel_abundance is numeric
df$rel_abundance <- as.numeric(df$rel_abundance)

# Check raw range BEFORE clipping
range(df$rel_abundance, na.rm = TRUE)
sum(df$rel_abundance < 0, na.rm = TRUE)
sum(df$rel_abundance > 1, na.rm = TRUE)

# Beta regression requires 0 < y < 1
eps <- 1e-6
df <- df %>%
  mutate(
    rel_abundance = pmin(pmax(rel_abundance, eps), 1 - eps)
  )

# Check again AFTER clipping
range(df$rel_abundance, na.rm = TRUE)
sum(df$rel_abundance <= 0, na.rm = TRUE)
sum(df$rel_abundance >= 1, na.rm = TRUE)

# Rebuild factors AFTER cleaning
df$disease_status <- factor(df$disease_status, levels = c("Control", "AD-CAA"))
df$Scan_ID <- factor(df$Scan_ID)
df$celltype <- factor(df$celltype)
df$region <- factor(df$region)
df$pathology <- factor(df$pathology, levels = c("AmyloidFree", "Amyloid"))

# Now create AD-only object from the cleaned df
dat_ad <- df %>%
  filter(disease_status == "AD-CAA")
```

```{r}
# AD-CAA only, since you want amyloid effect within disease
mean_results <- list()
contrast_results <- list()

for (rg in levels(dat_ad$region)) {
  for (ct in levels(dat_ad$celltype)) {
    
    tmp <- dat_ad %>%
      filter(region == rg, celltype == ct) %>%
      filter(!is.na(rel_abundance)) %>%
      droplevels()
    
    if (nrow(tmp) < 5) next
    if (n_distinct(tmp$Scan_ID) < 2) next
    if (n_distinct(tmp$pathology) < 2) next
    
    fit <- try(
      glmmTMB(
        rel_abundance ~ pathology + (1 | Scan_ID),
        data = tmp,
        family = beta_family(link = "logit")
      ),
      silent = TRUE
    )
    
    if (inherits(fit, "try-error")) next
    
    emm <- emmeans(fit, ~ pathology, type = "response")
    
    emm_df <- as.data.frame(emm) %>%
      rename(mean_abundance = response) %>%
      mutate(celltype = ct, region = rg)
    
    mean_results[[paste(rg, ct, sep = "__")]] <- emm_df
    
    cont <- contrast(
      emm,
      method = list(Amyloid_effect = c(-1, 1))
    )
    
    cont_df <- as.data.frame(summary(cont, infer = TRUE)) %>%
      mutate(celltype = ct, region = rg)
    
    contrast_results[[paste(rg, ct, sep = "__")]] <- cont_df
  }
}
```

```{r}
#P value adjustment (by region)
means_df <- bind_rows(mean_results)

contrast_df <- bind_rows(contrast_results) %>%
  group_by(region) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()
```

```{r}
#Means HeatMap
#Pivot (so each regions becomes a column)
mean_heat_df <- means_df %>%
  mutate(region_path = paste(region, pathology, sep = "_")) %>%
  group_by(celltype, region_path) %>%
  summarise(mean_abundance = mean(mean_abundance), .groups = "drop") %>%
  pivot_wider(names_from = region_path, values_from = mean_abundance)

desired_mean_cols <- c(
  "Parenchyma_AmyloidFree", "Parenchyma_Amyloid",
  "Arteries_AmyloidFree",   "Arteries_Amyloid",
  "Capillaries_AmyloidFree","Capillaries_Amyloid"
)

desired_mean_cols <- desired_mean_cols[desired_mean_cols %in% colnames(mean_heat_df)]

mean_mat <- as.matrix(mean_heat_df[, desired_mean_cols, drop = FALSE])
rownames(mean_mat) <- mean_heat_df$celltype
```

```{r}
#Build Significance tables 
#Only mark Amyloid positive columns 
mean_sig_df <- contrast_df %>%
  select(celltype, region, p_adj) %>%
  mutate(
    region_path = paste(region, "Amyloid", sep = "_"),
    sig = case_when(
      is.na(p_adj) ~ "",
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE ~ ""
    )
  ) %>%
  select(celltype, region_path, sig) %>%
  pivot_wider(names_from = region_path, values_from = sig)

mean_sig_mat <- matrix(
  "",
  nrow = nrow(mean_mat),
  ncol = ncol(mean_mat),
  dimnames = dimnames(mean_mat)
)

for (col in intersect(colnames(mean_sig_df), colnames(mean_mat))) {
  idx <- match(rownames(mean_mat), mean_sig_df$celltype)
  mean_sig_mat[, col] <- mean_sig_df[[col]][idx]
}

```

```{r}
#Mean Heatmap
pheatmap(
  mean_mat,
  color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
  scale = "none",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  border_color = NA,
  fontsize_row = 8,
  fontsize_col = 10,
  display_numbers = mean_sig_mat,
  number_color = "black",
  main = "Model-estimated Mean Cell-Type Abundance\nAmyloidFree vs Amyloid Across Regions (AD-CAA)"
)
```

```{r}
#Contrast Heatmap
#Red -> Enriched in Amyloid
#Blue -> Depleted in Amyloid 
contrast_heat_df <- contrast_df %>%
  mutate(log2_OR = log2(odds.ratio)) %>%
  select(celltype, region, log2_OR, p_adj) %>%
  pivot_wider(names_from = region, values_from = c(log2_OR, p_adj), names_sep = "___")

desired_contrast_cols <- c(
  "log2_OR___Parenchyma",
  "log2_OR___Arteries",
  "log2_OR___Capillaries"
)

desired_contrast_cols <- desired_contrast_cols[desired_contrast_cols %in% colnames(contrast_heat_df)]

contrast_mat <- as.matrix(contrast_heat_df[, desired_contrast_cols, drop = FALSE])
rownames(contrast_mat) <- contrast_heat_df$celltype
colnames(contrast_mat) <- gsub("^log2_OR___", "", desired_contrast_cols)

# Columns for adjusted p-values
desired_padj_cols <- c(
  "p_adj___Parenchyma",
  "p_adj___Arteries",
  "p_adj___Capillaries"
)

desired_padj_cols <- desired_padj_cols[desired_padj_cols %in% colnames(contrast_heat_df)]

padj_mat <- as.matrix(contrast_heat_df[, desired_padj_cols, drop = FALSE])
rownames(padj_mat) <- contrast_heat_df$celltype
colnames(padj_mat) <- gsub("^p_adj___", "", desired_padj_cols)

# Significance labels
contrast_sig_mat <- matrix(
  "",
  nrow = nrow(padj_mat),
  ncol = ncol(padj_mat),
  dimnames = dimnames(padj_mat)
)

contrast_sig_mat[padj_mat < 0.05] <- "*"
contrast_sig_mat[padj_mat < 0.01] <- "**"
contrast_sig_mat[padj_mat < 0.001] <- "***"
contrast_sig_mat[is.na(padj_mat)] <- ""
```

```{r}
#Contrasts Heatmap
breaks <- seq(-2.5, 2.5, length.out = 101)

pheatmap(
  contrast_mat,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = breaks,
  scale = "none",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  border_color = NA,
  fontsize_row = 8,
  fontsize_col = 10,
  display_numbers = contrast_sig_mat,
  number_color = "black",
  main = "Amyloid Effect by Region\n(log2 Odds Ratio, Amyloid vs AmyloidFree)"
)
```

```{r}
amyloid_df <- contrast_df %>%
  mutate(
    log2_OR = log2(odds.ratio),
    type = "Amyloid"
  ) %>%
  select(celltype, region, log2_OR, p.value, p_adj, type)

```

# Disease Effect (All regions \~ only AmyloidFree)

library(dplyr)

```{r}
library(dplyr)

df <- read.csv("spatial_celltype_proportions.csv", stringsAsFactors = FALSE)

# Make sure rel_abundance is numeric
df$rel_abundance <- as.numeric(df$rel_abundance)

# Check raw range BEFORE clipping
range(df$rel_abundance, na.rm = TRUE)
sum(df$rel_abundance < 0, na.rm = TRUE)
sum(df$rel_abundance > 1, na.rm = TRUE)

# Beta regression requires 0 < y < 1
eps <- 1e-6
df <- df %>%
  mutate(
    rel_abundance = pmin(pmax(rel_abundance, eps), 1 - eps)
  )

# Check again AFTER clipping
range(df$rel_abundance, na.rm = TRUE)
sum(df$rel_abundance <= 0, na.rm = TRUE)
sum(df$rel_abundance >= 1, na.rm = TRUE)

# Rebuild factors AFTER cleaning
df$disease_status <- factor(df$disease_status, levels = c("Control", "AD-CAA"))
df$Scan_ID <- factor(df$Scan_ID)
df$celltype <- factor(df$celltype)
df$region <- factor(df$region)
df$pathology <- factor(df$pathology, levels = c("AmyloidFree", "Amyloid"))

# Now create AmyloidFree-only object from the cleaned df
dat_dis <- df %>%
  filter(pathology == "AmyloidFree")
```

```{r}
# Amyloid Free only, since you want disease effects between FDX
mean_results <- list()
contrast_results <- list()

for (rg in levels(dat_dis$region)) {
  for (ct in levels(dat_dis$celltype)) {
    
    tmp <- dat_dis %>%
      filter(region == rg, celltype == ct) %>%
      filter(!is.na(rel_abundance)) %>%
      droplevels()
    
    if (nrow(tmp) < 5) next
    if (n_distinct(tmp$Scan_ID) < 2) next
    if (n_distinct(tmp$disease_status) < 2) next
    
    fit <- try(
      glmmTMB(
        rel_abundance ~ disease_status + (1 | Scan_ID),
        data = tmp,
        family = beta_family(link = "logit")
      ),
      silent = TRUE
    )
    
    if (inherits(fit, "try-error")) next
    
    emm <- emmeans(fit, ~ disease_status, type = "response")
    
    emm_df <- as.data.frame(emm) %>%
      rename(mean_abundance = response) %>%
      mutate(celltype = ct, region = rg)
    
    mean_results[[paste(rg, ct, sep = "__")]] <- emm_df
    
    cont <- contrast(
      emm,
      method = list(Disease_effect = c(-1, 1))
    )
    
    cont_df <- as.data.frame(summary(cont, infer = TRUE)) %>%
      mutate(celltype = ct, region = rg)
    
    contrast_results[[paste(rg, ct, sep = "__")]] <- cont_df
  }
}
```

```{r}
#P value adjustment (by region)
means_df_dis <- bind_rows(mean_results)

contrast_df_dis <- bind_rows(contrast_results) %>%
  group_by(region) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()
```

```{r}
#Means HeatMap
#Pivot (so each regions becomes a column)
mean_heat_df_dis <- means_df_dis %>%
  mutate(region_status = paste(region, disease_status, sep = "_")) %>%
  group_by(celltype, region_status) %>%
  summarise(mean_abundance = mean(mean_abundance), .groups = "drop") %>%
  pivot_wider(names_from = region_status, values_from = mean_abundance)

desired_mean_cols_dis <- c(
  "Parenchyma_Control", "Parenchyma_AD-CAA",
  "Arteries_Control",   "Arteries_AD-CAA",
  "Capillaries_Control","Capillaries_AD-CAA"
)

desired_mean_cols_dis <- desired_mean_cols_dis[desired_mean_cols_dis %in% colnames(mean_heat_df_dis)]

mean_mat_dis <- as.matrix(mean_heat_df_dis[, desired_mean_cols_dis, drop = FALSE])
rownames(mean_mat_dis) <- mean_heat_df_dis$celltype
```

```{r}
#Build Significance tables 
#Only mark Amyloid positive columns 
mean_sig_df_dis<- contrast_df_dis %>%
  select(celltype, region, p_adj) %>%
  mutate(
    region_status = paste(region, "AD-CAA", sep = "_"),
    sig = case_when(
      is.na(p_adj) ~ "",
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE ~ ""
    )
  ) %>%
  select(celltype, region_status, sig) %>%
  pivot_wider(names_from = region_status, values_from = sig)

mean_sig_mat_dis <- matrix(
  "",
  nrow = nrow(mean_mat_dis),
  ncol = ncol(mean_mat_dis),
  dimnames = dimnames(mean_mat_dis)
)

for (col in intersect(colnames(mean_sig_df_dis), colnames(mean_mat_dis))) {
  idx <- match(rownames(mean_mat_dis), mean_sig_df_dis$celltype)
  mean_sig_mat_dis[, col] <- mean_sig_df_dis[[col]][idx]
}

```

```{r}
#Mean Heatmap
pheatmap(
  mean_mat_dis,
  color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
  scale = "none",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  border_color = NA,
  fontsize_row = 8,
  fontsize_col = 10,
  display_numbers = mean_sig_mat_dis,
  number_color = "black",
  main = "Model-estimated Mean Cell-Type Abundance\nControl vs AD-CAA Across Amyloid-Free Regions"
)
```

```{r}
#Contrast Heatmap
#Red -> Enriched in AD
#Blue -> Depleted in AD
contrast_heat_df <- contrast_df_dis %>%
  mutate(log2_OR = log2(odds.ratio)) %>%
  select(celltype, region, log2_OR, p_adj) %>%
  pivot_wider(names_from = region, values_from = c(log2_OR, p_adj), names_sep = "___")

desired_contrast_cols <- c(
  "log2_OR___Parenchyma",
  "log2_OR___Arteries",
  "log2_OR___Capillaries"
)

desired_contrast_cols <- desired_contrast_cols[desired_contrast_cols %in% colnames(contrast_heat_df)]

contrast_mat <- as.matrix(contrast_heat_df[, desired_contrast_cols, drop = FALSE])
rownames(contrast_mat) <- contrast_heat_df$celltype
colnames(contrast_mat) <- gsub("^log2_OR___", "", desired_contrast_cols)

# Columns for adjusted p-values
desired_padj_cols <- c(
  "p_adj___Parenchyma",
  "p_adj___Arteries",
  "p_adj___Capillaries"
)

desired_padj_cols <- desired_padj_cols[desired_padj_cols %in% colnames(contrast_heat_df)]

padj_mat <- as.matrix(contrast_heat_df[, desired_padj_cols, drop = FALSE])
rownames(padj_mat) <- contrast_heat_df$celltype
colnames(padj_mat) <- gsub("^p_adj___", "", desired_padj_cols)

# Significance labels
contrast_sig_mat <- matrix(
  "",
  nrow = nrow(padj_mat),
  ncol = ncol(padj_mat),
  dimnames = dimnames(padj_mat)
)

contrast_sig_mat[padj_mat < 0.05] <- "*"
contrast_sig_mat[padj_mat < 0.01] <- "**"
contrast_sig_mat[padj_mat < 0.001] <- "***"
contrast_sig_mat[is.na(padj_mat)] <- ""
```

```{r}
#Contrasts Heatmap
breaks <- seq(-2.5, 2.5, length.out = 101)

pheatmap(
  contrast_mat,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = breaks,
  scale = "none",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  border_color = NA,
  fontsize_row = 8,
  fontsize_col = 10,
  display_numbers = contrast_sig_mat,
  number_color = "black",
  main = "Disease Effect (AD-CAA vs Control)\nlog2 Odds Ratio, AmyloidFree ROI"
)
```

```{r}
disease_df <- contrast_df_dis %>%
  mutate(
    log2_OR = log2(odds.ratio),
    type = "Disease"
  ) %>%
  select(celltype, region, log2_OR, p.value, p_adj, type)
```

# Overall Effect

```{r}
overall_dat <- df %>%
  mutate(
    group = case_when(
      disease_status == "Control" ~ "Control_AmyloidFree",
      disease_status == "AD-CAA" & pathology == "AmyloidFree" ~ "AD_AmyloidFree",
      disease_status == "AD-CAA" & pathology == "Amyloid" ~ "AD_Amyloid",
      TRUE ~ NA_character_
    ),
    group = factor(group, levels = c("Control_AmyloidFree", "AD_AmyloidFree", "AD_Amyloid"))
  ) %>%
  filter(!is.na(group))

mean_results_overall <- list()
contrast_results_overall <- list()
weight_df <- list()

for (rg in levels(overall_dat$region)) {
  
  # region-specific amyloid weight within AD
  w <- overall_dat %>%
    filter(region == rg, disease_status == "AD-CAA") %>%
    summarise(w = mean(pathology == "Amyloid")) %>%
    pull(w)
  
  weight_df[[rg]] <- data.frame(region = rg, w = w)
  
  for (ct in levels(overall_dat$celltype)) {
    
    tmp <- overall_dat %>%
      filter(region == rg, celltype == ct) %>%
      filter(!is.na(rel_abundance), is.finite(rel_abundance)) %>%
      droplevels()
    
    if (nrow(tmp) < 5) next
    if (n_distinct(tmp$Scan_ID) < 2) next
    if (n_distinct(tmp$group) < 2) next
    
    fit <- try(
      glmmTMB(
        rel_abundance ~ group + (1 | Scan_ID),
        data = tmp,
        family = beta_family(link = "logit")
      ),
      silent = TRUE
    )
    
    if (inherits(fit, "try-error")) next
    
    # Means for the 3 observed groups
    emm <- try(
      emmeans(fit, ~ group, type = "response"),
      silent = TRUE
    )
    
    if (inherits(emm, "try-error")) next
    
    emm_df <- as.data.frame(summary(emm))
    value_col <- intersect(c("response", "prob", "emmean"), colnames(emm_df))[1]
    if (is.na(value_col)) next
    
    emm_df <- emm_df %>%
      rename(mean_abundance = all_of(value_col)) %>%
      mutate(celltype = ct, region = rg, weight = w)
    
    mean_results_overall[[paste(rg, ct, sep = "__")]] <- emm_df
    
    # Contrasts on link scale
    em_link <- emmeans(fit, ~ group)
    
    contr <- try(
      contrast(
        em_link,
        method = list(
          Disease_effect        = c(-1,  1,  0),
          Amyloid_effect        = c( 0, -1,  1),
          AD_overall_vs_Control = c(-1, (1 - w), w)
        )
      ),
      silent = TRUE
    )
    
    if (inherits(contr, "try-error")) next
    
    cont_df <- as.data.frame(summary(contr, infer = TRUE)) %>%
      mutate(celltype = ct, region = rg, weight = w)
    
    contrast_results_overall[[paste(rg, ct, sep = "__")]] <- cont_df
  }
}

means_df_overall <- bind_rows(mean_results_overall)
contrast_df_overall <- bind_rows(contrast_results_overall) %>%
  group_by(region, contrast) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()

weights_used <- bind_rows(weight_df)
weights_used
```

```{r}
# Extract the three group means
mean_wide_overall <- means_df_overall %>%
  select(celltype, region, group, mean_abundance, weight) %>%
  pivot_wider(names_from = group, values_from = mean_abundance)

# Compute weighted AD overall mean
mean_wide_overall <- mean_wide_overall %>%
  mutate(
    AD_overall = (1 - weight) * AD_AmyloidFree + weight * AD_Amyloid,
    Control_overall = Control_AmyloidFree
  )
```

```{r}
mean_heat_df_overall <- mean_wide_overall %>%
  select(celltype, region, Control_overall, AD_overall) %>%
  pivot_longer(
    cols = c(Control_overall, AD_overall),
    names_to = "status",
    values_to = "mean_abundance"
  ) %>%
  mutate(
    status = recode(status,
                    "Control_overall" = "Control",
                    "AD_overall" = "AD_overall"),
    region_status = paste(region, status, sep = "_")
  ) %>%
  select(celltype, region_status, mean_abundance) %>%
  pivot_wider(names_from = region_status, values_from = mean_abundance)
```

```{r}
desired_mean_cols_overall <- c(
  "Parenchyma_Control", "Parenchyma_AD_overall",
  "Arteries_Control",   "Arteries_AD_overall",
  "Capillaries_Control","Capillaries_AD_overall"
)

desired_mean_cols_overall <- desired_mean_cols_overall[
  desired_mean_cols_overall %in% colnames(mean_heat_df_overall)
]

mean_mat_overall <- as.matrix(mean_heat_df_overall[, desired_mean_cols_overall, drop = FALSE])
rownames(mean_mat_overall) <- mean_heat_df_overall$celltype

mean_sig_df_overall <- contrast_df_overall %>%
  filter(contrast == "AD_overall_vs_Control") %>%
  select(celltype, region, p_adj) %>%
  mutate(
    region_status = paste(region, "AD_overall", sep = "_"),
    sig = case_when(
      is.na(p_adj) ~ "",
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE ~ ""
    )
  ) %>%
  select(celltype, region_status, sig) %>%
  pivot_wider(names_from = region_status, values_from = sig)

mean_sig_mat_overall <- matrix(
  "",
  nrow = nrow(mean_mat_overall),
  ncol = ncol(mean_mat_overall),
  dimnames = dimnames(mean_mat_overall)
)

for (col in intersect(colnames(mean_sig_df_overall), colnames(mean_mat_overall))) {
  idx <- match(rownames(mean_mat_overall), mean_sig_df_overall$celltype)
  mean_sig_mat_overall[, col] <- mean_sig_df_overall[[col]][idx]
}
```

```{r}
pheatmap(
  mean_mat_overall,
  color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
  scale = "none",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  border_color = NA,
  fontsize_row = 8,
  fontsize_col = 10,
  display_numbers = mean_sig_mat_overall,
  number_color = "black",
  main = "Model-estimated Mean Cell-Type Abundance\nControl vs Weighted AD Overall"
)
```

```{r}
contrast_heat_df_overall <- contrast_df_overall %>%
  filter(contrast == "AD_overall_vs_Control") %>%
  select(celltype, region, estimate, p_adj) %>%
  pivot_wider(names_from = region, values_from = c(estimate, p_adj), names_sep = "___")

desired_contrast_cols_overall <- c(
  "estimate___Parenchyma",
  "estimate___Arteries",
  "estimate___Capillaries"
)

desired_contrast_cols_overall <- desired_contrast_cols_overall[
  desired_contrast_cols_overall %in% colnames(contrast_heat_df_overall)
]

contrast_mat_overall <- as.matrix(
  contrast_heat_df_overall[, desired_contrast_cols_overall, drop = FALSE]
)

rownames(contrast_mat_overall) <- contrast_heat_df_overall$celltype
colnames(contrast_mat_overall) <- gsub("^estimate___", "", desired_contrast_cols_overall)

desired_padj_cols_overall <- c(
  "p_adj___Parenchyma",
  "p_adj___Arteries",
  "p_adj___Capillaries"
)

desired_padj_cols_overall <- desired_padj_cols_overall[
  desired_padj_cols_overall %in% colnames(contrast_heat_df_overall)
]

padj_mat_overall <- as.matrix(
  contrast_heat_df_overall[, desired_padj_cols_overall, drop = FALSE]
)

rownames(padj_mat_overall) <- contrast_heat_df_overall$celltype
colnames(padj_mat_overall) <- gsub("^p_adj___", "", desired_padj_cols_overall)

contrast_sig_mat_overall <- matrix(
  "",
  nrow = nrow(padj_mat_overall),
  ncol = ncol(padj_mat_overall),
  dimnames = dimnames(padj_mat_overall)
)

contrast_sig_mat_overall[padj_mat_overall < 0.05] <- "*"
contrast_sig_mat_overall[padj_mat_overall < 0.01] <- "**"
contrast_sig_mat_overall[padj_mat_overall < 0.001] <- "***"
contrast_sig_mat_overall[is.na(padj_mat_overall)] <- ""
```

```{r}
#Contrasts Heatmap
breaks <- seq(-2.5, 2.5, length.out = 101)

pheatmap(
  contrast_mat_overall,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = breaks,
  scale = "none",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  border_color = NA,
  fontsize_row = 8,
  fontsize_col = 10,
  display_numbers = contrast_sig_mat_overall,
  number_color = "black",
  main = "Overall AD vs Control Effect by Region\n(Weighted AD Composition)"
)
```

```{r}
overall_df <- contrast_df_overall %>%
  filter(contrast == "AD_overall_vs_Control") %>%
  mutate(
    log2_OR = estimate / log(2),
    type = "Overall"
  ) %>%
  select(celltype, region, log2_OR, p.value, p_adj, type)
```

# Maximal Pathology Comparison

```{r}
comparison_dat <- df %>%
  mutate(
    group2 = case_when(
      disease_status == "Control" & pathology == "AmyloidFree" ~ "Control_AmyloidFree",
      disease_status == "AD-CAA" & pathology == "Amyloid" ~ "AD_Amyloid",
      TRUE ~ NA_character_
    ),
    group2 = factor(group2, levels = c("Control_AmyloidFree", "AD_Amyloid"))
  ) %>%
  filter(!is.na(group2))

mean_results_comp <- list()
contrast_results_comp <- list()
fit_status_comp <- list()

for (rg in levels(comparison_dat$region)) {
  for (ct in levels(comparison_dat$celltype)) {
    
    tmp <- comparison_dat %>%
      filter(region == rg, celltype == ct) %>%
      filter(!is.na(rel_abundance), is.finite(rel_abundance)) %>%
      droplevels()
    
    status <- data.frame(
      region = rg,
      celltype = ct,
      n_rows = nrow(tmp),
      n_scans = n_distinct(tmp$Scan_ID),
      n_groups = n_distinct(tmp$group2),
      fit_ok = FALSE,
      converged = NA,
      pd_hessian = NA,
      stringsAsFactors = FALSE
    )
    
    if (nrow(tmp) < 5) {
      fit_status_comp[[paste(rg, ct, sep = "__")]] <- status
      next
    }
    if (n_distinct(tmp$Scan_ID) < 2) {
      fit_status_comp[[paste(rg, ct, sep = "__")]] <- status
      next
    }
    if (n_distinct(tmp$group2) < 2) {
      fit_status_comp[[paste(rg, ct, sep = "__")]] <- status
      next
    }
    
    fit <- try(
      glmmTMB(
        rel_abundance ~ group2 + (1 | Scan_ID),
        data = tmp,
        family = beta_family(link = "logit")
      ),
      silent = TRUE
    )
    
    if (inherits(fit, "try-error")) {
      fit_status_comp[[paste(rg, ct, sep = "__")]] <- status
      next
    }
    
    status$fit_ok <- TRUE
    
    # Extract convergence diagnostics if available
    s <- try(summary(fit), silent = TRUE)
    if (!inherits(s, "try-error")) {
      status$converged <- isTRUE(fit$sdr$pdHess)
      status$pd_hessian <- isTRUE(fit$sdr$pdHess)
    }
    
    emm <- try(
      emmeans(fit, ~ group2, type = "response"),
      silent = TRUE
    )
    
    if (inherits(emm, "try-error")) {
      fit_status_comp[[paste(rg, ct, sep = "__")]] <- status
      next
    }
    
    emm_df <- as.data.frame(summary(emm))
    value_col <- intersect(c("response", "prob", "emmean"), colnames(emm_df))[1]
    if (is.na(value_col)) {
      fit_status_comp[[paste(rg, ct, sep = "__")]] <- status
      next
    }
    
    emm_df <- emm_df %>%
      rename(mean_abundance = all_of(value_col)) %>%
      mutate(celltype = ct, region = rg)
    
    mean_results_comp[[paste(rg, ct, sep = "__")]] <- emm_df
    
    em_link <- emmeans(fit, ~ group2)
    
    contr <- try(
      contrast(
        em_link,
        method = list(
          AD_Amyloid_vs_Control = c(-1, 1)
        )
      ),
      silent = TRUE
    )
    
    if (inherits(contr, "try-error")) {
      fit_status_comp[[paste(rg, ct, sep = "__")]] <- status
      next
    }
    
    cont_df <- as.data.frame(summary(contr, infer = TRUE)) %>%
      mutate(celltype = ct, region = rg)
    
    contrast_results_comp[[paste(rg, ct, sep = "__")]] <- cont_df
    fit_status_comp[[paste(rg, ct, sep = "__")]] <- status
  }
}

means_df_comp <- bind_rows(mean_results_comp)

contrast_df_comp <- bind_rows(contrast_results_comp) %>%
  group_by(region) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()

fit_status_comp_df <- bind_rows(fit_status_comp)
```

```{r}
table(fit_status_comp_df$region, fit_status_comp_df$pd_hessian, useNA = "ifany")
fit_status_comp_df %>% filter(pd_hessian == FALSE | is.na(pd_hessian))
```

```{r}
mean_heat_df_comp <- means_df_comp %>%
  mutate(region_group = paste(region, group2, sep = "_")) %>%
  group_by(celltype, region_group) %>%
  summarise(mean_abundance = mean(mean_abundance), .groups = "drop") %>%
  pivot_wider(names_from = region_group, values_from = mean_abundance)

desired_mean_cols_comp <- c(
  "Parenchyma_Control_AmyloidFree", "Parenchyma_AD_Amyloid",
  "Arteries_Control_AmyloidFree",   "Arteries_AD_Amyloid",
  "Capillaries_Control_AmyloidFree","Capillaries_AD_Amyloid"
)

desired_mean_cols_comp <- desired_mean_cols_comp[
  desired_mean_cols_comp %in% colnames(mean_heat_df_comp)
]

mean_mat_comp <- as.matrix(mean_heat_df_comp[, desired_mean_cols_comp, drop = FALSE])
rownames(mean_mat_comp) <- mean_heat_df_comp$celltype

mean_sig_df_comp <- contrast_df_comp %>%
  select(celltype, region, p_adj) %>%
  mutate(
    region_group = paste(region, "AD_Amyloid", sep = "_"),
    sig = case_when(
      is.na(p_adj) ~ "",
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE ~ ""
    )
  ) %>%
  select(celltype, region_group, sig) %>%
  pivot_wider(names_from = region_group, values_from = sig)

mean_sig_mat_comp <- matrix(
  "",
  nrow = nrow(mean_mat_comp),
  ncol = ncol(mean_mat_comp),
  dimnames = dimnames(mean_mat_comp)
)

for (col in intersect(colnames(mean_sig_df_comp), colnames(mean_mat_comp))) {
  idx <- match(rownames(mean_mat_comp), mean_sig_df_comp$celltype)
  mean_sig_mat_comp[, col] <- mean_sig_df_comp[[col]][idx]
}
```

```{r}
pheatmap(
  mean_mat_comp,
  color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
  scale = "none",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  border_color = NA,
  fontsize_row = 8,
  fontsize_col = 10,
  display_numbers = mean_sig_mat_comp,
  number_color = "black",
  main = "Model-estimated Mean Cell-Type Abundance\nControl Amyloid-Free vs AD Amyloid"
)
```

```{r}
contrast_heat_df_comp <- contrast_df_comp %>%
  select(celltype, region, estimate, p_adj) %>%
  pivot_wider(names_from = region, values_from = c(estimate, p_adj), names_sep = "___")

desired_contrast_cols_comp <- c(
  "estimate___Parenchyma",
  "estimate___Arteries",
  "estimate___Capillaries"
)

desired_contrast_cols_comp <- desired_contrast_cols_comp[
  desired_contrast_cols_comp %in% colnames(contrast_heat_df_comp)
]

contrast_mat_comp <- as.matrix(
  contrast_heat_df_comp[, desired_contrast_cols_comp, drop = FALSE]
)

rownames(contrast_mat_comp) <- contrast_heat_df_comp$celltype
colnames(contrast_mat_comp) <- gsub("^estimate___", "", desired_contrast_cols_comp)

desired_padj_cols_comp <- c(
  "p_adj___Parenchyma",
  "p_adj___Arteries",
  "p_adj___Capillaries"
)

desired_padj_cols_comp <- desired_padj_cols_comp[
  desired_padj_cols_comp %in% colnames(contrast_heat_df_comp)
]

padj_mat_comp <- as.matrix(
  contrast_heat_df_comp[, desired_padj_cols_comp, drop = FALSE]
)

rownames(padj_mat_comp) <- contrast_heat_df_comp$celltype
colnames(padj_mat_comp) <- gsub("^p_adj___", "", desired_padj_cols_comp)

contrast_sig_mat_comp <- matrix(
  "",
  nrow = nrow(padj_mat_comp),
  ncol = ncol(padj_mat_comp),
  dimnames = dimnames(padj_mat_comp)
)

contrast_sig_mat_comp[padj_mat_comp < 0.05] <- "*"
contrast_sig_mat_comp[padj_mat_comp < 0.01] <- "**"
contrast_sig_mat_comp[padj_mat_comp < 0.001] <- "***"
contrast_sig_mat_comp[is.na(padj_mat_comp)] <- ""
```

```{r}
#Contrasts Heatmap
breaks <- seq(-2.5, 2.5, length.out = 101)

pheatmap(
  contrast_mat_comp,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = breaks,
  scale = "none",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  border_color = NA,
  fontsize_row = 8,
  fontsize_col = 10,
  display_numbers = contrast_sig_mat_comp,
  number_color = "black",
  main = "AD Amyloid vs Control Amyloid-Free by Region"
)
```

```{r}
maxpath_df <- contrast_df_comp %>%
  mutate(
    log2_OR = estimate / log(2),
    type = "MaxPathology"
  ) %>%
  select(celltype, region, log2_OR, p.value, p_adj, type)
```

# Heat Maps by Region

```{r}
amyloid_df_clean <- contrast_df %>%
  transmute(
    celltype,
    region,
    type = "Amyloid",
    log2_OR = log2(odds.ratio),
    p_adj
  ) %>%
  distinct()

disease_df_clean <- contrast_df_dis %>%
  transmute(
    celltype,
    region,
    type = "Disease",
    log2_OR = log2(odds.ratio),
    p_adj
  ) %>%
  distinct()

overall_df_clean <- contrast_df_overall %>%
  filter(contrast == "AD_overall_vs_Control") %>%
  transmute(
    celltype,
    region,
    type = "Overall",
    log2_OR = estimate / log(2),
    p_adj
  ) %>%
  distinct()

maxpath_df_clean <- contrast_df_comp %>%
  transmute(
    celltype,
    region,
    type = "MaxPathology",
    log2_OR = estimate / log(2),
    p_adj
  ) %>%
  distinct()

combined_df_clean <- bind_rows(
  amyloid_df_clean,
  disease_df_clean,
  overall_df_clean,
  maxpath_df_clean
)
```

```{r}
library(dplyr)
library(tidyr)
library(pheatmap)

regions <- c("Arteries", "Capillaries", "Parenchyma")

for (rg in regions) {
  
  tmp <- combined_df_clean %>%
    filter(region == rg) %>%
    group_by(celltype, type) %>%
    summarise(
      log2_OR = first(log2_OR),
      p_adj   = first(p_adj),
      .groups = "drop"
    )
  
  # -----------------------------
  # Effect matrix
  # -----------------------------
  heat_df <- tmp %>%
    select(celltype, type, log2_OR) %>%
    pivot_wider(names_from = type, values_from = log2_OR)
  
  desired_cols <- c("Amyloid", "Disease", "Overall", "MaxPathology")
  desired_cols <- desired_cols[desired_cols %in% colnames(heat_df)]
  
  mat <- as.matrix(heat_df[, desired_cols, drop = FALSE])
  rownames(mat) <- heat_df$celltype
  
  # -----------------------------
  # Significance matrix
  # -----------------------------
  sig_df <- tmp %>%
    mutate(sig = case_when(
      is.na(p_adj) ~ "",
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE ~ ""
    )) %>%
    select(celltype, type, sig) %>%
    pivot_wider(names_from = type, values_from = sig)
  
  sig_df[is.na(sig_df)] <- ""
  
  sig_mat <- as.matrix(sig_df[, desired_cols, drop = FALSE])
  rownames(sig_mat) <- sig_df$celltype
  
  # -----------------------------
  # Plot
  # -----------------------------
  mat_display <- pmax(pmin(mat, 2.5), -2.5)
  mat_cluster <- mat_display
  mat_cluster[!is.finite(mat_cluster)] <- 0
  
  pheatmap(
    mat_display,
    color = colorRampPalette(c("blue", "white", "red"))(100),
    breaks = seq(-2.5, 2.5, length.out = 101),
    scale = "none",
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    clustering_distance_rows = dist(mat_cluster),
    border_color = NA,
    fontsize_row = 7,
    fontsize_col = 10,
    display_numbers = sig_mat,
    number_color = "black",
    na_col = "grey90",
    main = paste("Cell-Type Effects Across Contrasts —", rg)
  )
}
```

```{r}
```

```{r}

```

## Supplementary

```{r}
#effect magnitude similarity across analyses
combined_df %>%
  ggplot(aes(x = type, y = abs(log2_OR))) +
  geom_boxplot() +
  facet_wrap(~region) +
  theme_classic() +
  labs(y = "|log2 OR|", title = "Effect Size Distribution by Contrast")
```

```{r}
#Effect size vs Significance 
combined_df %>%
  ggplot(aes(x = log2_OR, y = -log10(p.value))) +
  geom_point() +
  facet_grid(region ~ type) +
  theme_classic()
```

```{r}
#% of significant cell types 
combined_df %>%
  group_by(type, region) %>%
  summarise(
    n_sig = sum(p_adj < 0.05, na.rm = TRUE),
    n_tested = n(),
    .groups = "drop"
  )
combined_df %>%
  group_by(type, region) %>%
  summarise(sig = mean(p_adj < 0.05), .groups = "drop") %>%
  ggplot(aes(x = type, y = sig)) +
  geom_col() +
  facet_wrap(~region) +
  theme_classic() +
  labs(y = "% significant")
```

```{r}
combined_df %>%
  ggplot(aes(x = type, y = SE)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 1.2) +
  facet_wrap(~region) +
  theme_classic() +
  labs(
    x = "",
    y = "Standard error",
    title = "Standard error comparison across contrast types"
  )
```

```{r}
amyloid_se <- contrast_df %>%
  mutate(type = "Amyloid") %>%
  select(celltype, region, SE, type)

disease_se <- contrast_df_dis %>%
  mutate(type = "Disease") %>%
  select(celltype, region, SE, type)

overall_se <- contrast_df_overall %>%
  filter(contrast == "AD_overall_vs_Control") %>%
  mutate(type = "Overall") %>%
  select(celltype, region, SE, type)

maxpath_se <- contrast_df_comp %>%
  mutate(type = "MaxPathology") %>%
  select(celltype, region, SE, type)

combined_se <- bind_rows(
  amyloid_se,
  disease_se,
  overall_se,
  maxpath_se
)
```

```{r}
#Standard Error Distributions
combined_se %>%
  ggplot(aes(x = type, y = SE)) +
  geom_boxplot() +
  facet_wrap(~region) +
  theme_classic() +
  labs(title = "Standard Error Distribution by Contrast")
```

```{r}
library(ggplot2)
library(dplyr)

combined_df %>%
  left_join(combined_se, by = c("celltype", "region", "type")) %>%
  mutate(sig = p_adj < 0.05) %>%
  ggplot(aes(x = log2_OR, y = SE)) +
  
  # Points colored by significance
  geom_point(aes(color = sig), size = 2, alpha = 0.8) +
  
  # Smooth trend
  geom_smooth(method = "loess", se = FALSE, color = "red", linewidth = 0.7) +
  
  # Global SE reference
  geom_hline(
    yintercept = median(combined_se$SE, na.rm = TRUE),
    linetype = "dashed",
    color = "black"
  ) +
  
  # Faceting
  facet_grid(region ~ type) +
  
  # Colors
  scale_color_manual(
    values = c("FALSE" = "grey70", "TRUE" = "#D7191C"),
    name = "FDR < 0.05"
  ) +
  
  # Labels
  labs(
    x = "Effect size (log2 Odds Ratio)",
    y = "Standard Error (SE)",
    title = "Effect Size vs Uncertainty Across Contrast Strategies"
  ) +
  
  theme_classic(base_size = 13) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.position = "top"
  )
```

## Identify Robust Cell types

```{r}
robust_df <- combined_df %>%
  mutate(sig = p_adj < 0.05) %>%
  group_by(celltype) %>%
  summarise(
    n_sig = sum(sig, na.rm = TRUE),
    mean_effect = mean(log2_OR, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(n_sig))

robust_cts <- robust_df %>%
  filter(n_sig >= 3)  # across contrasts
```

```{r}
robust_df %>%
  ggplot(aes(x = reorder(celltype, n_sig), y = n_sig)) +
  geom_col() +
  coord_flip() +
  theme_classic() +
  labs(title = "Robust Cell Types Across Contrasts")
```

## Cluster Cells types by response pattern

```{r}
cluster_mat <- combined_df %>%
  select(celltype, region, type, log2_OR) %>%
  mutate(feature = paste(type, region, sep = "_")) %>%
  select(celltype, feature, log2_OR) %>%
  tidyr::pivot_wider(names_from = feature, values_from = log2_OR)

mat <- as.matrix(cluster_mat[, -1])
rownames(mat) <- cluster_mat$celltype

mat[is.na(mat)] <- 0
```

```{r}
pheatmap(
  mat,
  scale = "row",
  clustering_distance_rows = "euclidean",
  clustering_method = "ward.D2",
  main = "Cell-Type Response Patterns Across Contrasts"
)
```

```{r}
abs_mat_df <- combined_df %>%
  select(celltype, region, type, log2_OR) %>%
  mutate(feature = paste(type, region, sep = "_")) %>%
  select(celltype, feature, log2_OR) %>%
  tidyr::pivot_wider(names_from = feature, values_from = log2_OR)

abs_mat <- as.matrix(abs_mat_df[, -1])
rownames(abs_mat) <- abs_mat_df$celltype

# Replace NA (missing fits) with 0 for visualization
abs_mat[is.na(abs_mat)] <- 0

max_val <- max(abs(abs_mat), na.rm = TRUE)

breaks <- seq(-max_val, max_val, length.out = 100)
colors <- colorRampPalette(c("blue", "white", "red"))(99)

pheatmap(
  abs_mat,
  color = colors,
  breaks = breaks,
  scale = "none",  # <-- CRITICAL
  clustering_distance_rows = "euclidean",
  clustering_method = "ward.D2",
  border_color = NA,
  fontsize_row = 7,
  fontsize_col = 9,
  main = "Absolute Effect Sizes Across Contrasts\n(log2 Odds Ratio)"
)
```
